---
title: "Building a Rigsarkivet Information Package for Statistical Data using R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building a Rigsarkivet Information Package for Statistical Data using R}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

# Introduction

This tool is designed to emit an statistical Information Package
described by [Annex 9 of Executive Order 128](https://en.rigsarkivet.dk/wp-content/uploads/2022/09/Executive-Order-on-Information-Packages128.pdf)
suitable for testing with ASTA and then submission to the
Rigsarkivet. 

THIS PACKAGE IS UNOFFICIAL AND NOT ENDORSED BY THE RIGSARKIVET!

Prerequisites:

* A contextDocumentationIndex.xml file
* An archiveIndex.xml file
* A data package identifier (FD.XXXXX)

All of the above are supplied by the Rigsarkivet when you are invited
to submit a data package.

If you wish to test this package, the example FD.18005 data package
can be used as a basis for recreation (and is the basis of the
examples in this package
 
# Installation

The R package has been written for R version 3.6.0 and above.

It is not currently available on CRAN - please see the Development
section for install instructions

## Development

```{r eval=FALSE}
library(devtools)
install_github("BiomathematicsAndStatisticsScotland/rigsarkivet_r_infopkg/RigsArkivetRInfoPkg")
```

To build the package you will need KnitR and Rmarkdown packages
installed. For subsequent use, base R will suffice. 

# Basic usage, emitting files for use with ASTA

ASTA expects to take one or more datasets exported in a specific
format and guides the user through the process of combining them into
an information package. 

You can export your dataset(s) in this way as follows, using the
fd_18005_r example dataframe shipped with the package:

```{r eval=FALSE}
library(RigsArkivetRInfoPkg)
process_as_asta_export_script(fd_18005_r, 
                              "my_dataset",
                              output_dir=tempdir(),
                              factors_to_codes=TRUE)
```

This will output 3, possibly four files:

my_dataset.csv
:a CSV file containing the data
 
my_dataset_VARIABEL.txt
:The variable list (with types and KODELISTE references)
 
my_dataset_VARIABELBESKRIVELSE.txt
:The descriptions of each variable
 
my_dataset_KODELISTE.txt
:the code list for all categorical/factor variables in the dataset. Only produced if there are categorical/factor variables and factors_to_codes = TRUE

# Basic usage, single R data frame

It is also possible to export a complete data package, ready for
testing with ASTA. Here's a simple example for a single data frame,
using data contained in the RigsArkivetRInfoPkg package:

```{r eval=FALSE}
    library(RigsArkivetRInfoPkg)

    pkg_output_dir=tempdir()

    pkg_description = list(
     archive_index=system.file("extdata", "FD_18005_archiveIndex.xml", package="RigsArkivetRInfoPkg"),
     context_doc_index=system.file("extdata", "FD_18005_contextDocumentationIndex.xml", package="RigsArkivetRInfoPkg"),
     pkg_id="FD.18005",
                   tables=list(
                       list(
                          name="R_testfil",
                          label_file=fd_18005_r_labels,                             
                          key_variable=c("child_id"),
                          description="Danish Longitudinal Research Study of Grandparents, Parents and Children - this is data 1",
                          table_dataset=fd_18005_r)))
    process_full_info_pkg(pkg_description,
                          output_dir = pkg_output_dir)   
```

This will emit the entire data package into the directory
pkg_output_dir. Note that there will be some warnings about context
documentation when doing this as I do not ship all the contents of
FD.18005. See "Handling Context Documentation" later in this
document for a discussion of the issues here. 

You can then take the emitted package and test it in ASTA. 

# Advanced usage, multiple R data frames linked by key variables

For non-trivial datasets you will have multiple datasets linked by
variable references. The Information Package specification supports this

```{r eval=FALSE}
    library(RigsArkivetRInfoPkg)

    pkg_output_dir=tempdir()

    pkg_description = list(
    archive_index=system.file("extdata", "FD_18005_archiveIndex.xml", package="RigsArkivetRInfoPkg"),
    context_doc_index=system.file("extdata", "FD_18005_contextDocumentationIndex.xml", package="RigsArkivetRInfoPkg"),
    pkg_id="FD.18005",
    tables=list(
        list(
            name="R_testfil",
            label_file=fd_18005_r_labels,                             
            key_variable=c("child_id"),
            description="Danish Longitudinal Research Study of Grandparents, Parents and Children - this is data 1",
            table_dataset=fd_18005_r,
            reference=list(
                list(
                    other_dataset="childrens_pets",
                    other_key_variable="child_id",
                    our_key_variable="child_id")
            )
        ),
        list(
            name="childrens_pets",
            label_file=example_table2_labels,
            key_variable=c("pet_id"),
            description="Childrens' pets, to provide a simple example of a 2nd table",
            table_dataset=example_table2)))

    process_full_info_pkg(pkg_description,
                          output_dir = pkg_output_dir)
```

# Use of KODELISTE

The option factors_to_codes determines whether a KODELISTE section is
emitted. If you choose to emit KODELISTE:

* All factor variables will be converted to their numeric (integer)
  equivalents
* The textual representation will be used to generate the KODELISTE 

# Use of "User codes"/BRUGERKODE

It is possible to specify user codes for fields in your data that have
specific values with specific meaning (e.g. different kinds of missing
data). A simple example is given below, which indicates that certain
codes in each of the fields in our example dataset actually indicate
missing data.

NOTE: these user codes are DIFFERENT to the original FD.18005
information package due to coercion of these variables to an R factor
(which doesn't preserve code values and ignores levels which do not
exist in the data)

```{r eval=FALSE}
pkg_output_dir=file.path(tempdir(), "FD.18005")

pkg_info=list(
    archive_index=system.file("extdata", "FD_18005_archiveIndex.xml", package="RigsArkivetRInfoPkg"),
    context_doc_index=system.file("extdata", "FD_18005_contextDocumentationIndex.xml", package="RigsArkivetRInfoPkg"),
    pkg_id="FD.18005",
                  tables=list(
                      list(
                          name="R_testfil",
                          label_file=fd_18005_r_labels,
                          user_codes=list( gender='3',
                                           grade='3',
                                           education='6',
                                           hobby=c('8')
                                          ),                          
                          key_variable=c("child_id"),
                          description="Danish Longitudinal Research Study of Grandparents, Parents and Children - this is data 1 ",
                          table_dataset=fd_18005_r))
)


process_full_info_pkg(pkg_info,
                      output_dir = pkg_output_dir)               
```

In general, if the data is originated from R and contains factors,
using factors_to_codes=TRUE is likely to be easier. 

# Handling Context Documentation

Generating suitably-formatted TIFF files for contextDocumentation is
beyond the scope of this package and you are advised to consult the
Danish National Archives copious documentation on this topic. For many
use-cases the ImageMagick "convert" tool can be used, similar to the following:

```{sh eval=FALSE}
  convert -background white -alpha remove -depth 8 -compress lzw\ 
          -density 300 \
          source_file.pdf contextDocumentation/docCollection1/1/1.tif
```


However the process_full_info_package routine will check to see if a
single contextDocumentation/docCollection1 folder exists, and that it
contains subdirectories and a single TIF file for each <document>
entry in contextDocumentationIndex.xml

If it doesn't find them, a warning will be issued

# Limitations and observations

* The code on which this package is based has special handling of
  attributes added by the "haven" package when used to represent data
  types in SAS, SPSS and STATA in datasets that were subsequently
  exported to R. While this code has been preserved, it is untested
  at present as I had no suitable datasets to use as a test case. 

* Interaction with the tidyverse (specifically tibbles) is untested,
  but no problems are anticipated

* Full tests and documentation are NOT YET COMPLETE

